Variable Density Tesseroids: Gravity fields calculation in spherical coordinates using variable densities in depth

### 2017-08-09 manuscript/manuscript.tex
Small changes in manuscript

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/tesseroid-uieda.pdf}
\caption{
A spherical prism, also known as  Tesseroid, in a geocentric spherical coordinate system, with a computation point $P$ and its local north oriented Cartesian coordinate system. Picture made by \citep{Uieda2015}.
}
\label{fig:tesseroid-uieda}
\end{figure}

@@

It's well known that the GLQ method becomes less accurate when the computation point is closer to the tesseroid \citep{Ku1977} or for smaller GLQ order. 
Although the model developed by \citet{Uieda2016} uses a second order GLQ, it ensures the accuracy implementing a modified version of the adaptive discretization of \citet{Li2011}.
It consist in splitting the tesseroid into smaller ones when a certain distance-size ratio ($D$) is greater than a predefined value that controls the precision of the computation.  \citet{Uieda2016} have also objectively obtained standard values of $D$ for gravity potential, gradients and tensor components comparing the fields generated by a spherical shell, which constitutes a special case of the volume integrals that has analytical solution \citep{LaFehr1991, Mikuska2006, Grombein2013}\todo{citar mas??}.
The modifications made by \citet{Uieda2016} to the original adaptive discretization method \citep{Li2011} consist in:
(1) a faster calculation of the distance from the computation point to the tesseroid and 
(2) the application of a stack based algorithm that speeds up the computation and gives more control over the recursion step.

@@

In our case, we will suppose that the tesseroid has an arbitrary variable density in depth, this means that only depends on the radius spherical coordinate. Thus, the integrals for the gravity fields are slightly modified.

@@

From equation \ref{eq:glq-var-dens} we can observe that the GLQ approximates the gravity fields of a tesseroid with variable density as the ones generated by point masses located at the nodes of the Legendre polynomials.
Furthermore, the information of the density function is summarised as the values it assumes in those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with a well fitted discretization algorithm, approximates the gravity fields with good accuracy and precision.\todo{no se si poner esta ultima oracion}

@@

In case of variable density tesseroids, the comparison must obviously be made with a variable density spherical shell, whose analytical solution will depend on the chosen density function.
Because the literature has not offered it yet, we needed to develop ourselves.

Lets consider a spherical shell with inner and outer radii $R_1$ and $R_2$, respectivetly, which density depends on the radius spherical coordinate (Figure \ref{fig:spherical-shell}).
The gravity potential generated by the shell on an arbitrary external point $Q(0,0,r)$ located along the $z$ axe at a distance $r$ can be written as follows:

@@

For any density $\rho(r')$, the gradient and Marussi tensor components can be obtained as \citet{Grombein2013} does. Only the vertical gradient component ($g_z$) and the diagonal Marussi tensor components ($g_{xx}$, $g_{yy}$, $g_{zz}$) are not zero, and can be easily determined from the potential $V_\text{sh}$:

@@

For practical purposes we are going to obtain the gravity potential for two very different functions in the perspective of they variation in $r'$: a linear density and an exponential density.

@@

If the density of the spherical shell is a linear function in the radius coordinate, i.e.:

@@

Although it's a Python library, the core functions of the computation are written in Cython, a superset of Python language that allows us generate efficient C code through the Cython compiler.
This produces a precompiled library whose functions and classes can be called from regular Python code, what allowed us to make it compatible with the preexisting open-source library Fatiando a Terra \citep{Uieda2013}.

@@

Instead of increasing the GLQ orders, what leads to a more time consuming computation, adaptive discretization algorithms have been applied in previous works \citep{Li2011, Uieda2016}.
They basically consist in dividing the tesseroid based on a ratio between its size and the distance between its geometric centre and the computation point.

In order to make the computation more efficient, we use only second order GLQ for each integration, and we ensure the accuracy of the method through the application of the modified adaptive discretization algorithm developed by \citet{Uieda2016}.

Given an arbitrary tesseroid, \citet{Uieda2016} redefined its dimensions: $L_\lambda$, $L_\phi$ and $L_r$, where the former ones are the arc-distances measured along its top surface and $L_r$ is the difference between the radii of the top and bottom surfaces.
We must check that the following inequality is satisfied:

@@

In summary, the distance-size ratio $D$ determines indirectly how many times the tesseroids will be divided and therefore both the accuracy of the algorithm and its computation time.
In this way, the modified version of the adaptive discretization algorithm efficiently improves its accuracy and gives the user more control over the computation: it allows to stop the division of the tesseroids when their dimensions are lower than a certain value (e.g. less than 10cm), which does not modifies significantly the desired gravitational field.

@@

In order to test the new variable density tesseroid code and to estimate the error of the approximation we have compared the numerical model with the analytical solutions for a spherical shell with variable density for two special cases: a linear and an exponential density function.
This error estimations are also useful in the determination of the value for the distance-size ratio $D$, such the difference between the numerical model and the analytical solutions are below an acceptable threshold.

@@

We computed the three gravity fields for different values of $D$, ranging from 0.5 to 10 with a step of 0.5 on every point of these four grids.
Then, we calculated the maximum difference between these results and the value of the analytical solution at the same height.

@@

\begin{table}
\caption{
    Description of the synthetic grids where the comparison of the numerical model against the analytical solutions for a spherical shell was done. This collection of grids tries to include every possible scenario on which the numerical model can present a different behaviour depending on the size of the grid, its location or its height. Every grid consists in a set of 10$\times$10 points.
}
\label{tab:grids}
\begin{tabular}{lcccc}
    Grid & Size & Lat. extension & Lon. extension & Height \\ \hline
    Pole & $1^\circ \times 1^\circ$ & $89^\circ - 90^\circ$ & $0^\circ - 1^\circ$ & 2km \\
    Equator & $1^\circ \times 1^\circ$ & $0^\circ - 1^\circ$ & $0^\circ - 1^\circ$ & 2km \\
    Satellite & $1^\circ \times 1^\circ$ & $89^\circ - 90^\circ$ & $0^\circ - 1^\circ$ & 260km \\
    Big Grid & $30^\circ \times 30^\circ$ & $60^\circ - 90^\circ$ & $0^\circ - 30^\circ$ & 2km \\
\end{tabular}
\end{table}

@@

As a simple case, we proposed a linear density function such as the spherical shell has a density of 2670~kg/m$^3$ on the surface of the Earth, and 3300~kg/m$^3$ at 35 km of depth:

@@

We can see that for a linear density, the calculation of the potential and the $g_z$ component of its gradient needs a value of $D=1$ and $D=2$, respectively, in order to ensure an accuracy of 10$^{-1}\%$.
On the other hand, the $g_{zz}$ computations show a noticeable difference between the thin and the thick shell: the first one needs a value of $D=8$ while the later only a $D=3.5$.
We will keep the more conservative result of $D=8$ in case of computing the $g_{zz}$ tensor component in case of a linear density tesseroid.

>>>>

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{figures/tesseroid-uieda.pdf}
\caption{
A spherical prism, also known as  Tesseroid, in a geocentric spherical coordinate system, with a computation point $P$ and its local north oriented Cartesian coordinate system. Picture made by \citet{Uieda2015}.
}
\label{fig:tesseroid-uieda}
\end{figure}

@@

It's well known that the GLQ method becomes less accurate when the computation point is closer to the tesseroid \citep{Ku1977} or for smaller GLQ order. 
Although the model developed by \citet{Uieda2016} uses a second order GLQ, it ensures the accuracy implementing a modified version of the adaptive discretization of \citet{Li2011}.
It consist in splitting the tesseroid into smaller ones when a certain distance-size ratio is greater than a predefined value ($D$) that controls the accuracy of the computation. 
\citet{Uieda2016} have also objectively obtained standard values of $D$ for gravity potential, gradients and tensor components comparing the fields generated by a spherical shell, which constitutes a special case of the volume integrals that has analytical solution \citep{LaFehr1991, Mikuska2006, Grombein2013}\todo{citar mas??}.
The modifications made by \citet{Uieda2016} to the original adaptive discretization method \citep{Li2011} consist in:
(1) a faster calculation of the distance from the computation point to the tesseroid and 
(2) the application of a stack based algorithm that speeds up the computation and gives more control over the recursion step.

@@

In our case, we will suppose that the tesseroid has an arbitrary variable density in depth, i.e. only depends on the radius spherical coordinate.
Thus, the integrals for the gravity fields are slightly modified:

@@

From equation \ref{eq:glq-var-dens} we can observe that the GLQ approximates the gravity fields of a tesseroid with variable density as the ones generated by point masses located at the nodes of the Legendre polynomials.
Furthermore, the information of the density function is summarised as the values it assumes on those nodes.
Although this may sound discouraging, we intent to prove otherwise and show that this method, along with a well fitted discretization algorithm, approximates the gravity fields with good accuracy and precision.\todo{no se si poner esta ultima oracion}

@@

In case of variable density tesseroids, the comparison must obviously be made with a variable density spherical shell, whose analytical solution will depend on the chosen density function.
Because the literature has not offered it yet, we needed to develop it ourselves.

Lets consider a spherical shell with inner and outer radii $R_1$ and $R_2$, respectivetly, which density depends on the radius spherical coordinate (Figure \ref{fig:spherical-shell}).
The gravity potential generated by the shell on an arbitrary external point $Q(0,0,r)$, located along the $z$ axe at a distance $r$ from the origin, can be written as follows:

@@

For any density $\rho(r')$, the gradient and Marussi tensor components can be obtained as \citet{Grombein2013} does.
Only the vertical gradient component ($g_z$) and the diagonal Marussi tensor components ($g_{xx}$, $g_{yy}$, $g_{zz}$) are non zero, and can be easily determined from the potential $V_\text{sh}$:

@@

For practical purposes we are going to obtain the gravity potential for two very different functions in the perspective of their variation in $r'$: a linear density and an exponential density.

@@

If the density of the spherical shell is a linear function on the radius coordinate, i.e.:

@@

Although it's a Python library, the core functions of the computation are written in Cython, a superset of Python language that allows us to generate efficient C code through the Cython compiler.
This produces a precompiled library whose functions and classes can be called from regular Python code, what allowed us to make it compatible with the preexisting open-source library Fatiando a Terra \citep{Uieda2013}.

@@

Instead of increasing the GLQ orders to improve the accuracy, what leads to a more time consuming computation, adaptive discretization algorithms have been applied in previous works \citep{Li2011, Uieda2016}.
They basically consist in dividing the tesseroid based on a ratio between its size and the distance between its geometric centre and the computation point.

In order to make the computation more efficient, we use only second order GLQ for each integration, and we ensure the accuracy of the method through the application of the modified adaptive discretization algorithm developed by \citet{Uieda2016}.
Given an arbitrary tesseroid, \citet{Uieda2016} redefined its dimensions: $L_\lambda$, $L_\phi$ and $L_r$, where the former ones are the arc-distances measured along its top surface and $L_r$ is the difference between the radii of the top and bottom surfaces.
We must check that the following inequality is satisfied:

@@

In summary, the distance-size ratio $D$ determines indirectly how many times the tesseroids will be divided and therefore both the accuracy of the algorithm and its computation time.
In this way, the modified version of the adaptive discretization algorithm efficiently improves its accuracy and gives the user more control over the computation: it allows to stop the division of the tesseroids when their dimensions are lower than a certain value (e.g. less than 10 cm), which does not modifies significantly the desired gravitational field.

@@

In order to test the new variable density tesseroid code and to estimate the error of the approximation we have compared the numerical model with the analytical solutions for a spherical shell with variable density for two special cases: a linear and an exponential density function.
This error estimations are also useful in the determination of the minimum value for the distance-size ratio $D$ such the difference between the numerical model and the analytical solutions are below an acceptable threshold.

@@

We computed the three gravity fields for different values of $D$, ranging from 0.5 to 10 with a step of 0.5, on every point of these four grids.
Then, we calculated the maximum difference between these results and the value of the analytical solution at the same height.

@@

\begin{table}
\caption{
    Description of the synthetic grids on which the comparison of the numerical model against the analytical solutions for a spherical shell was done. This collection of grids tries to include every possible scenario on which the numerical model can present a different behaviour depending on the size of the grid, its location or its height. Every grid consists in a set of 10$\times$10 points.
}
\label{tab:grids}
\begin{tabular}{lcccc}
    Grid & Size & Lat. extension & Lon. extension & Height \\ \hline
    Pole & $1^\circ \times 1^\circ$ & $89^\circ - 90^\circ$ & $0^\circ - 1^\circ$ & 2km \\
    Equator & $1^\circ \times 1^\circ$ & $0^\circ - 1^\circ$ & $0^\circ - 1^\circ$ & 2km \\
    Satellite & $1^\circ \times 1^\circ$ & $89^\circ - 90^\circ$ & $0^\circ - 1^\circ$ & 260km \\
    Big Grid & $30^\circ \times 30^\circ$ & $60^\circ - 90^\circ$ & $0^\circ - 30^\circ$ & 2km \\
\end{tabular}
\end{table}

@@

As a simple case, we proposed a linear density function in such a way that the spherical shell has a density of 2670~kg/m$^3$ on the surface of the Earth, and 3300~kg/m$^3$ at 35 km of depth:

@@

For the linear density case, we can observe that the calculation of the potential and the $g_z$ component of its gradient needs a value of $D=1$ and $D=2$, respectively, in order to ensure an accuracy of 10$^{-1}\%$.
On the other hand, the $g_{zz}$ computations show a noticeable difference between the thin and the thick shell: the first one needs a value of $D=8$ while the later only a $D=3.5$.
We will keep the more conservative result of $D=8$ in case of computing the $g_{zz}$ tensor component in case of a linear density tesseroid.
